---
title: "Untitled"
author: "Amelia McNamara"
date: "3/11/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r cars}
## Working with geographic data

```{r}
#install.packages("ggmap")
library(ggmap)
```

## 
```{r}
str(crime) # crime data from ggmap package
library(mosaic)
```

## 
```{r}
midlat <- mean(~lat, data=crime, na.rm=TRUE)
midlon <- mean(~lon, data=crime, na.rm=TRUE)
m1 <- get_map(source='stamen', location=c(midlon, midlat), 
              maptype="watercolor", zoom=11)
p4 <- ggmap(m1)
p4 
```

## With points
```{r}
p4 + geom_point(aes(x=lon, y=lat), data=crime)
```

##
```{r}
p4 + 
  coord_cartesian() +
  geom_hex(aes(x=lon, y=lat), data=crime, alpha=0.8) + 
  guides(fill=guide_legend(alpha=0.5, title="Number of crimes")) + 
  xlab("") + ylab("") + theme(axis.line=element_blank(), 
                              axis.text=element_blank(), 
                              axis.ticks=element_blank(),
                              axis.title=element_blank())
```


```{r, echo=FALSE}
p4 <- p4 + 
  coord_cartesian() +
  geom_hex(aes(x=lon, y=lat), data=crime, alpha=0.8) + 
  guides(fill=guide_legend(alpha=0.5, title="Number of crimes")) + 
  xlab("") + ylab("") + theme(axis.line=element_blank(), 
                              axis.text.x=element_blank(), 
                              axis.text.y=element_blank(), 
                              axis.ticks=element_blank(),
                              axis.title.x=element_blank(), 
                              axis.title.y=element_blank())
```

```{r, eval=FALSE}
ggsave("beautiful_hex_map.jpg", plot=p4)
```