---
title: "Data Appendix"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading data

```{r, message=FALSE, warning=FALSE}
# install.packages("readr")
library(readr)
library(mosaic)
collegedata <- read_csv("~/collegedata.csv")
```


## Structure and names

```{r}
str(collegedata)
```

There are 10 variables in the data, and 81 observations. The variables are:

- an unnamed variable that seems to just be row numbers
- `rank` contains the rank of schools (integer)
- `score` contains the score of each school (integer)
- `name` is a `chr` string of each school's name.
- `location` is a `chr` string of the city and state of each school
- `tuitionAndFees` is a `chr`, including dollar signs and commas. More useful would be as a numeric variable with dollar amounts
- `totalEnrollment` is an `int`
- `fall2013AcceptanceRate` is a numeric variable
- `averageFreshmanRetentionRate` is a numeric variable
- `sixYearGraduation rate` is a numeric variable

Other than `tuitionAndFees` the variable types make sense to me. All of these variable names are pretty self-explanatory, although I might want to rename the first column or remove it altogether. 

## Variable analysis

```{r}
favstats(~rank, data=collegedata)
```
Minimum rank is 1, max is 75, no missing data. This all makes sense.

```{r}
favstats(~score, data=collegedata)
```
Minimum is 49 (need to look in to what this means) maximum is 100 (makes sense!). No missing data.

```{r, eval=FALSE}
tally(~name, data=collegedata)
```
Tried this and it is just a bunch of 1s and school names. 


**Etcetera**

## Most pressing data cleaning issues

- Need to work on removing dollar signs and commas from `tuitionAndFees`. 
- Need to look into whata score of 49 is and whether it makes sense. 
- Might want to split out just the state names from `location` so we can use state as a variable. 
