---
title: "Reproducible Research"
author: "Amelia McNamara"
date: "3/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reproducible research

- Karl Broman on [reproducible research](http://kbroman.org/steps2rr/)

## RMarkdown
- [Markdown syntax](http://daringfireball.net/projects/markdown/syntax)
- [knitr options](https://yihui.name/knitr/options/)
- .Rmd is RMarkdown, .Rnw is R and LaTeX

- You can use LaTeX
\[
\frac{var(\epsilon_i)}{var({y_i})}
\]

- [Bookdown](https://bookdown.org/yihui/bookdown/)
- To knit to a different format,

```{r, eval=FALSE}
?knit
```

- To pull out code,

```{r, eval=FALSE}
?purl
```

## git 
- git can check the difference between two text/code files
- track changes in those files
- commandline tool
- can use locally without GitHub
- [Happy git with R](http://happygitwithr.com/)

## GitHub 
- a way to share code
- can use Upload button to share code without commandline tool of git
+ [Free private GitHub repos for education](https://education.github.com/)